SFJ: An implementation of Semantic Featherweight Java

### 2020-04-23 conference/camera ready/paper.tex
on type system for SFJ

    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java as these types can be inhabited by assigning \emph{null} to all fields, but are not useful in practice.

@@

If \autoref{alg:types} returns $True$, meaning all type definitions in an SFJ program are valid, we can proceed with building the semantic subtyping.
Leveraging the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and define $Universe$ to be the set of all types.
The mapping for basic types is defined as follows.

@@

Some comments on \autoref{alg:subtyping} will follow.
In function $generateRelation$ we iterate over the set of classes in an SFJ program.
If the class currently being processed contains types in its fields or methods not present in the subtyping relation (line 5, 30, 42), then we add the current class to the list of \emph{unprocessed} classes (line 6) so we can process its later as it is currently undecidable.
The set of unprocessed classes will then be inspected again in a recursive call (line 10).
The next two functions of the algorithm, \emph{addClass} and \emph{checkSuperSet}, check subtyping for the current $class$ being processed and update $relation$, which is a mapping from a type to its subtypes and originally only consists of entries from \autoref{eq:initialSubtype}.
In function $addClass(class)$ we check if the type $class$ is a subtype of an existing type in $relation$ (lines 15-18), as well as the opposite, meaning if $class$ is a supertype of an existing type in $relation$ (line 19).
In order to do so $checkSuperSet$ checks all fields (lines 28-39) and all methods (lines 40-51) in $class$ and compares them with an \emph{existingClass} in $relation$.
If a subtyping relation is established, then it is added to $relation$ (line 53).
Finally, upon returning from $checkSuperSet$, we also add $class$ to its own relation (line 21) to satisfy reflexivity and to $Universe$ (line 22), which is a supertype of all types.

@@

It is worth noticing that the subtyping algorithm finds all nominal and structural subtypes of a given type.
This is due to the fact that all pairs of types are inspected.
Recall from \autoref{sec:intro} that nominal subtyping is name-based and given by the class hierarchy defined by the programmer, whereas structural subtyping is structure-based and given by the set-inclusion of fields and methods.
In particular, it is contra-variant with respect to this set-inclusion.
\autoref{alg:subtyping} finds all structural subtypes of a given class because it checks that its fields and methods are a superset of existing types in $relation$.
On the other hand, it also finds all nominal subtypes because a class inherits all fields and methods of its superclass and as such its fields and methods are a superset of its superclass.
This means that checking for structural subtyping is enough because nominal subtyping will be captured due to inheritance of fields and methods.

Finally, a note on complexity.
The complexity of \autoref{alg:types} is $\mathcal{O}(n)$, and the complexity of \autoref{alg:subtyping} is $\mathcal{O}(n^{2})$, with $n$ being the size of the input.
The reason for a quadratic complexity \autoref{alg:subtyping} is due to the symmetric check of structural subtyping between a $class$ and an $existingClass$ in $relation$.
Notice that if we were to only work with nominal subtyping, then we only require traversing the class hierarchy once,  which gives an $\mathcal{O}(n)$ complexity.

@@

The subtyping relation generated by \autoref{alg:subtyping} is used in the type system for SFJ terms, given in \autoref{sec:terms}.
The complete type system and its properties are presented in relevant work \cite{Dardha2013,Dardha2017} and are beyond the scope of this paper.
Let us briefly comment on typing expressions in SFJ.
Field access $e.a$, method call $e.m(e)$, variables $x$ and object creation $\synnew C(\wt{e})$ are typed in the same way as in Java.

For the constants $c$, these are the literal values in our language such as $42$, $3.14$ or $True$.
In most languages such as Java, literal values are cast to the type that is expected in the context they are used.
For example, a method $\textbf{int}\ getNum()\{return\ 42\}$ automatically casts the value to the type $int$.

However, in our system, types represent sets of values which must interpret boolean connectives in a set-theoretic way.
Given the boolean type $int\ \textbf{and not}\ byte$, this type represents the values $-2^31$ to $2^31 - 1$ from which are removed the values $-128$ to $127$.
If we were to type the value $42$ as an integer, this would be accepted by the boolean type, but would not be in the set of values of the type, thus breaking our semantic type system.
Therefore, for all constants $c$, we use a strong typing to assign values the most restrictive types possible, i.e. the type with the smallest amount of values which still contains this value.
With this typing rule, the value $42$ is assigned the type $Byte$, so would correctly not be accepted as a value of the type $(int\ \textbf{and not}\ byte)$.

We have now defined how to type any expression that can be formed by the syntax and so have defined our typing rules.

\subsection{Polygons: Continued}

>>>>

    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method computes and returns the length of its longest diagonal; otherwise, there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that these types can be defined and inhabited in Java by assigning \emph{null} to all fields in a class, however they are not useful in practice.

@@

If \autoref{alg:types} returns $True$, meaning all type definitions in a program are valid, we can then build the semantic subtyping.
Leveraging the interpretation of types as sets of values to define semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and map $Universe$ to the set of all types.
The full mapping for basic types is defined as follows:

@@

Some comments on \autoref{alg:subtyping} follow.
In function $generateRelation$ we iterate over the set of classes in an SFJ program.
If the class currently being processed contains types in its fields or methods not present in the subtyping relation (lines 5, 30, 42), then we add the current class to the list of \emph{unprocessed} classes (line 6) so we can process its fields and methods first and the class itself later after having all required type information.
The set of unprocessed classes will then be inspected again in a recursive call (line 10).
The next two functions of the algorithm, \emph{addClass} and \emph{checkSuperSet}, check subtyping for the current $class$ being processed and update $relation$, which is a mapping from a type to its subtypes and originally only consists of entries from \autoref{eq:initialSubtype}.
In function $addClass(class)$ we check if the type $class$ is a subtype of an existing type in $relation$ (lines 15-18), as well as the opposite, meaning if $class$ is a supertype of an existing type in $relation$ (line 19).
In order to do so $checkSuperSet$ checks all fields (lines 28-39) and all methods (lines 40-51) in $class$ and compares them with an \emph{existingClass} in $relation$.
If a subtyping relation is established, then it is added to $relation$ (line 53).
Finally, upon returning from $checkSuperSet$, we also add $class$ to its own relation (line 21) to satisfy reflexivity and to $Universe$ (line 22), which is a supertype of all types.

@@

It is worth noticing that the subtyping algorithm finds all nominal and structural subtypes of a given type. This is due to the fact that all pairs of types are inspected.
Recall from \autoref{sec:intro} that nominal subtyping is name-based and given by the class hierarchy defined by the programmer, whether structural subtyping is structure-based and given by the set-inclusion of fields and methods. In particular, it is contra-variant with respect to this set-inclusion.
\autoref{alg:subtyping} finds all structural subtypes of a given class because it checks that its fields and methods are a superset of existing types in $relation$. For example, all classes are structural subtypes of type $empty = []$.
On the other hand, it also finds all nominal subtypes because a class inherits all fields and methods of its superclass and as such its fields and methods are a superset of its superclass.
This means that checking for structural subtyping is enough because nominal subtyping will be captured due to inheritance of fields and methods.

Finally, a note on complexity.
The complexity of \autoref{alg:types} is $\mathcal{O}(n)$, and the complexity of \autoref{alg:subtyping} is $\mathcal{O}(n^{2})$, with $n$ being the size of the input.
The reason for a quadratic complexity of \autoref{alg:subtyping} is due to the symmetric check of structural subtyping between a $class$ and an $existingClass$ in $relation$.
Notice that if we were to only work with nominal subtyping, then we only require traversing the class hierarchy once,  which gives an $\mathcal{O}(n)$ complexity.

@@

The type system for the SFJ language, given in \autoref{sec:terms}, is based on the type system by Dardha \emph{et al.} \cite{Dardha2013,Dardha2017} where the formal typing rules and soundness properties are detailed. As these formal developments are beyond the scope of this paper, we discuss typing for SFJ only informally.

A program $(\tilde L,e)$ is well typed if both $\tilde{L}$ and $e$ are well typed.
Class declaration $L$ and method declaration $M$ are well typed if all their components are well typed.
Let us move onto expressions $E$.
Field access $e.a$, method call $e.m(e)$ and object creation $\synnew C(\wt{e})$ are typed in the same way as in Java: we inspect the type of the field and the type of the method and its arguments to determine the type of the field access and method call, respectively. The type of an object creation is determined by the type of its class.
Regarding constants, in order to respect the set-theoretic interpretation of types as sets of values, we type constants with the most restrictive type, i.e., the type representing the smallest set of values containing the value itself.
For example, the type system would assign to the value $42$ the type \textbf{byte}, which is the smallest in the sequence \textbf{byte}, \textbf{short}, \textbf{int} (see \autoref{eq:initialSubtype} for details).

% Let us move onto constants, where we can also illustrate the boolean connectives. 
% Typing constants in Java uses type casting where the type of a constant changes to be adapted to the context it is being used in.
% %For example, a method $\textbf{long}\ getNum()\{return\ 42;\}$ casts $42$ to \textbf{long}.
% %Instead, in variable type inference $\textbf{var}\ x = 42$, the compiler treats $42$ as having type \textbf{int}, as this is the most efficient type for hardware operations.

% Since in our framework types represent sets of values and the boolean connectives \textbf{and}, \textbf{not}, \textbf{or} are interpreted in a set-theoretic way, typing constants changes from Java as follows.
% Consider the boolean type \textbf{int and} (\textbf{not byte}), this type is interpreted as the set of values $-2^31$ to $2^31 - 1$ removing the values $-128$ to $127$.
% In SFJ, if we were to type $42$ as \textbf{int}, this would also mean that $42$ can be typed by \textbf{int and} (\textbf{not byte}), however, the value itself would not be in the set of values of the type, thus breaking our semantic type system.
% Therefore, our design choice is to type constants with the most restrictive type, i.e., the type representing the smallest set of values containing the value itself.
% With this typing rule, the value $42$ has type \textbf{byte}, and as a consequence it cannot have type \textbf{int and} (\textbf{not byte}).

Finally, the subtyping relation generated by \autoref{alg:subtyping} is used in the type system for the SFJ language via a subsumption typing rule:
 $$
 \frac{\Gamma\vdash e:\alpha_1 \qquad \alpha_1 \leq \alpha_2}{\Gamma \vdash e:\alpha_2}
 $$
We read this typing rule as follows:
if an expression $e$ is of type $\alpha_1$ under a typing context $\Gamma$ (details of a typing context are irrelevant here) and type $\alpha$ is a subtype $\leq$ of $\alpha_2$, then expression $e$ can be typed with $\alpha_2$.

\subsection{Polygons: continued}
